\documentclass[11pt,a4paper,oneside,numbers=endperiod]{scrartcl}
% Source: http://www.cs.princeton.edu/courses/archive/fall03/cs528/handouts/fibonacci%20heaps.pdf

\usepackage{hyperref}    % Hyperlinking and PDF metadata
\usepackage{polyglossia} % Babel for XeLaTex
\usepackage{microtype}   % Microtypography improvements
\usepackage{fontspec}    % Custom font selection
\usepackage{amsmath,amsthm} % Math goodies
\usepackage{titling}     % Custom \maketitle format
\usepackage{paralist}    % More list environments
\usepackage{csquotes}    % Context-sensitive quotations
\usepackage{subcaption}  % Captions for subfigures
\usepackage{tikz}        % Diagram drawing
\usepackage{comment}     % Comments ignored during rendering

% -----------------------------------------------------------------------------------
% Fonts
% -----------------------------------------------------------------------------------
\usepackage{mathpazo}
\setmainfont{Linux Libertine}
\setsansfont{Linux Biolinum}

% -----------------------------------------------------------------------------------
% Metadata and bibliography
% -----------------------------------------------------------------------------------
\setmainlanguage{english} % Polyglossia language
\hypersetup{
    unicode,
    colorlinks,
    linkcolor=purple,
    urlcolor=purple,
    citecolor=purple,
    pdftitle=Fibonacci Heaps and Their Uses in Improved Network Optimization %
        Algorithms,
    pdfauthor={Michael L. Fredman, Robert Endre Tarjan}}
\usepackage[backend=biber,autolang=hyphen,sortlocale=en_US,style=numeric-comp,
    sorting=nyt]{biblatex}
\addbibresource{references.bib}
% Make the References section text smaller
\renewcommand*{\bibfont}{\small}

% -----------------------------------------------------------------------------------
% Convenient shortcuts
% -----------------------------------------------------------------------------------
\newcommand*\bigO[1]{\ensuremath{\mathcal{O}(#1)}}
\newcommand*\term[1]{\emph{#1}}
\newcommand*\oper[1]{\textsc{#1}}
\newcommand*\opargs[2]{\textsc{#1}(\ensuremath{#2})}

\newenvironment{step}[1]
    {\vskip .8em plus .4em minus .3em%
        \textit{#1.}\hskip .8em}
    {\vskip .8em plus .4em minus .3em}

\newtheorem{theorem}{Theorem}
\newtheorem{lemma}{Lemma}
\newtheorem{corollary}{Corollary}
\providecommand*{\lemmaautorefname}{Lemma}
\providecommand*{\corollaryautorefname}{Corollary}

\theoremstyle{remark}
\newtheorem*{remark}{Remark}

% -----------------------------------------------------------------------------------
% Title format
% -----------------------------------------------------------------------------------
%\def\maketitlehooka{\begin{flushleft}\addtolength{\leftskip}{3em}}
\pretitle{\centering\LARGE\sffamily\bfseries}
\posttitle{\par\vskip 1em}
%\preauthor{\large\bfseries}
%\postauthor{\par\vskip 0.3em}
\predate{}
\postdate{}
%\def\maketitlehookd{\end{flushleft}}

% -----------------------------------------------------------------------------------
% TikZ/PGF settings
% -----------------------------------------------------------------------------------
\tikzset{sibling distance=3em,level distance=3em,%
	nodes={draw,circle,font=\sffamily\small,%
	minimum size=1.6em,inner sep=0}}

\begin{document}
\title{Fibonacci Heaps and Their Uses in Improved Network Optimization Algorithms}
\author{Michael L. Fredman\thanks{University of California, San Diego, La Jolla,
    California}
\and Robert Endre Tarjan\thanks{AT\&T Bell Laboratories, Murray Hill, New Jersey}}
\date{}
\maketitle

\begin{abstract}
    In this paper we develop a new data structure for implementing heaps (priority
    queues). Our structure, \term{Fibonacci heaps} (abbreviated \term{F-heaps}),
    extends the binomial queues proposed by Vuillemin and studied further by Brown.
    F-heaps support arbitrary deletion from an \(n\)-item heap in \bigO{\log n}
    amortized time and all other standard heap operations in \bigO{1} amortized time.
    Using F-heaps we are able to obtain improved running times for several network
    optimization algorithms. In particular, we obtain the following worst-case
    bounds, where \(n\) is the number of vertices and \(m\) the number of edges in
    the problem graph:
    \begin{enumerate}
        \item \bigO{n \log n + m} for the single-source shortest path problem with
            nonnegative edge lengths, improved from \bigO{m \log_{(m/n + 2)} n};
        \item \bigO{n^2 \log n + n m} for the all-pairs shortest path problem,
            improved from \bigO{n m \log_{(m/n + 2)} n};
        \item \bigO{n^2 \log n + n m} for the assignment problem (weighted bipartite
            matching), improved from \bigO{n m \log_{(m/n + 2)} n};
        \item \bigO{m \beta(m, n)} for the minimum spanning tree problem, improved
            from \bigO{m \log\log_{(m/n + 2)} n}, where \(\beta(m, n) = \min \{
            i \mid \log^{(i)} n \leq m/n \}\). Note that \(\beta(m, n) \leq \log^*
            n\) if \(m \geq n\).
    \end{enumerate}

    One of these results, the improved bound for minimum spanning trees is the most
    striking, although all the results give asymptotic improvements for graphs of
    appropriate densities.
\end{abstract}

\begin{comment}
Categories and Subject Descriptors:
    E.1 [\textbf{Data}]: Data Structures---\textit{trees}; \textit{graphs};
    F.2.2 [\textbf{Analysis of Algorithms and Problem Complexity}]: Nonnumerical
        Algorithms and Problems---\textit{computations on discrete structures};
        \textit{sorting and searching};
    G.2.2 [\textbf{Discrete Mathematics}]: Graph Theory---\textit{graph algorithms};
        \textit{network problems}; \textit{trees}

General Terms: Algorithms, Theory

Additional Key Words and Phrases: Heap, matching, minimum spanning tree, priority
queue, shortest path

A preliminary version of this paper appeared in the \emph{Proceedings of the 25th
IEEE Symposium on the Foundations of Computer Science} (Singer Island, Fla.,
Oct~24--26). IEEE, New York, 1984, pp.~338--346, \textcopyright~IEEE. Any portion of
this paper that appeared in the preliminary version is reprinted with permission.

This research was supported in part by the National Science Foundation under grant
MCS~82-04031.

Permission to copy without fee all or part of this material is granted provided that
the copies are not made or distributed for direct commercial advantage, the ACM
copyright notice and the title of the publication and its date appear, and notice is
given that copying is by permission of the Association for Computing Machinery. To
copy otherwise, or to republish, requires a fee and/or specific permission.

\textcopyright~1987 ACM 0004-5411/87/0700-0596 \$1.50

{\footnotesize Journal of the Association for Computing Machinery, Vol.~34, No.~3,
July 1987, Pages 596--615.}
\end{comment}

\section{Introduction}
\input{01-introduction}

\section{Fibonacci Heaps}
\label{sec:f-heaps}
\input{02-f-heaps}

\section{Variants of Fibonacci Heaps}
\label{sec:variants}
\input{03-variants}

\section{Shortest Paths}
\label{sec:shortest-paths}
\input{04-shortest-paths}

\section{Minimum Spanning Trees}
\label{sec:mst}
\input{05-mst}

\section{Recent Results and Open Problems}
\label{sec:results}
\input{06-results}

\paragraph{Acknowledgement.} We thank Dan Sleator for perceptive suggestions that
helped to simplify our data structure and Hal Gabow for pointing out the additional
applications of F-heaps mentioned in \autoref{sec:results}.

\printbibliography[title=References,heading=bibintoc]

\end{document}
% vim: set et sw=4 ts=4 sts=4:
